Reduced plasticity and variance in physiological rates of ectotherm populations under climate change

### 2024-03-13 ms.qmd
Re-organise and format for GCB/Ecol Letters

---
title: "Title: Reduced plasticity and variance in physiological rates of ectotherm populations under climate change"
bibliography: ./bib/refs.bib
csl: ./bib/science-advances.csl
format:
  docx:
    reference-doc: ./bib/template.docx
editor_options: 
  chunk_output_type: console
execute:
  freeze: auto  # re-render only when source changes
  cache: false
  echo: false
  warning: false
  error: false
  include: true
crossref:
  fig-title: '**Fig.**'
  fig-labels: arabic
  title-delim: "-"
  fig-prefix: "**Fig.**"
  tbl-prefix: "Table"
---

@@

## **Short Title**: Reduced physiological rate variation in ectotherms

## Authors:

@@

@@

## **Teasor**

Climate change reduces physiological rate variation in ectotherms to impact ecological and evolutionary processes.

@@

@@

Using a large database on physiological rates[@Seebacher2015-wi] for marine, freshwater and terrestrial ectotherms we apply new effect size estimates [@Noble2022-ty] that capture changes in mean physiological rates ($lnRR_{Q_{10}}$), and changes in their relative variance ($lnCVR_{Q_{10}}$), standardised to a 10&deg;C temperature difference. These standardised effects sizes can be converted to percentage differences to permit comparisons between the relative magnitudes of rate changes resulting from acclimation and relative variance changes at high temperatures. Acclimation responses can be determined by comparing acclimation $lnRR_{Q_{10}}$ ($lnRR_{Q_{10_{acclimation}}}$) to acute $lnRR_{Q_{10}}$ ($lnRR_{Q_{10_{acute}}}$) from ectotherms acclimated to different temperatures. The difference in $lnRR_{Q_{10_{acclimation}}}$ compared to $lnRR_{Q_{10_{acute}}}$ measures the extent of acclimation (i.e., plasticity).

@@

Using the ERA5 climate model, we extracted a 72-year period (1950-2022) of either surface air temperature (0.01&deg; resolution) for both terrestrial and freshwater taxa, or sea surface temperature for the marine taxa (at 0.25&deg; resolution). We used the historical temperature data for each geographical location to calculate a measure of thermal variability and predcitability (see *Methods*). Thermal variability (i.e., $CV$) and predictability experienced by a population in the past did not explain acclimation capacity or changes in physiological rate variance among terrestrial, marine or freshwater populations (@fig-4; @fig-5).

@@

### ***Potential consequences of reduced plasticity and variance in physiological rates across ectotherms***

@@

## **Methods & Protocols**
### *Literature collection*
We compiled literature on ectothermic animals that measured physiological rates (e.g., metabolic rate) at two or more temperatures after having been acclimated (or acclimatized) at these temperatures. We used data from a previous meta-analysis [@Seebacher2015-wi] and updated Seebacher *et al.* @Seebacher2015-wi's data by extracting data from suitable studies from our own searches that followed the same search protocol. More specifically, we performed a literature search using the Web of Science database for articles or proceedings papers published in English from 2013 to 2017 (the date after Seebacher *et al.* [@Seebacher2015-wi] searches were conducted) using the following topic search string: *"(acclimat* AND (therm* OR temp*) NOT (plant* OR tree* OR forest* OR fung* OR mammal* OR marsup* OR bird* OR human OR exercis* OR train* OR hypoxi*))"*. We further limited to the following research areas: Anatomy Morphology; Biodiversity Conservation; Biology; Ecology; Endocrinology Metabolism; Entomology; Evolutionary Biology; Marine Freshwater Biology; Physiology; Respiratory System, Reproductive Biology, Zoology.

@@

## Data Availability 
All data and code used to reproduce analyses can be found on GitHub at: https://github.com/daniel1noble/Q10_meta_analysis.

>>>>

---
title: "Reduced plasticity and variance in physiological rates of ectotherm populations under climate change"
bibliography: ./bib/refs.bib
csl: ./bib/ecology-letters.csl
format:
  docx:
    reference-doc: ./bib/template.docx
editor_options: 
  chunk_output_type: console
execute:
  freeze: auto  # re-render only when source changes
  cache: false
  echo: false
  warning: false
  error: false
  include: true
crossref:
  fig-title: 'Figure'
  fig-labels: arabic
  title-delim: "-"
  fig-prefix: "Figure"
  tbl-prefix: "Table"
---

@@

@@

**Running head**: Reduced physiological rate variation in ectotherms

@@

@@

## **Materials and Methods**
### *Literature collection*
We compiled literature on ectothermic animals that measured physiological rates (e.g., metabolic rate) at two or more temperatures after having been acclimated (or acclimatized) at these temperatures. We used data from a previous meta-analysis [@Seebacher2015-wi] and updated Seebacher *et al.* @Seebacher2015-wi's data by extracting data from suitable studies from our own searches that followed the same search protocol. More specifically, we performed a literature search using the Web of Science database for articles or proceedings papers published in English from 2013 to 2017 (the date after Seebacher *et al.* [@Seebacher2015-wi] searches were conducted) using the following topic search string: *"(acclimat* AND (therm* OR temp*) NOT (plant* OR tree* OR forest* OR fung* OR mammal* OR marsup* OR bird* OR human OR exercis* OR train* OR hypoxi*))"*. We further limited to the following research areas: Anatomy Morphology; Biodiversity Conservation; Biology; Ecology; Endocrinology Metabolism; Entomology; Evolutionary Biology; Marine Freshwater Biology; Physiology; Respiratory System, Reproductive Biology, Zoology.

@@

@@

Thermal variability (i.e., $CV$) and predictability experienced by a population in the past did not explain acclimation capacity or changes in physiological rate variance among terrestrial, marine or freshwater populations (@fig-4; @fig-5).

@@

### ***Consequences of reduced plasticity and variance in physiological rates across ectotherms***

@@

@@

## Data and code availability 
All data and code used to reproduce analyses can be found on GitHub at: https://github.com/daniel1noble/Q10_meta_analysis.

## Author contributions
Conceptualization: DWAN, FK, FS, SN; Methodology: DWAN, AB, FK, FS, SN; Investigation: DWAN, FK, FS, SN; Visualization: DWAN; Supervision: DWAN, SN, FS; Writing—original draft: DWAN; Writing—review & editing: DWAN, AB, FK, FS, SN.

## Conflict of interest
Authors declare that they have no competing interests.
